---
title: "Exploring the BRFSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
```

### Load data

Make sure your data and R Markdown files are in the same directory. When loaded
your data file will be called `brfss2013`. Delete this note when before you submit 
your work. 

```{r load-data}
load("brfss2013_truncate.RData")

dim(brfss2013_truncate)
dimnames(brfss2013_truncate)
```

```{r}
dim(brfss2013_truncate)
```

## Part 1: Data
According to the The Behavioral Risk Factor Surveillance System (BRFSS) website, data is collected from 50 states and 3 US territories in USA via telephone. Surveys are done on adults 18 years-olds.  
Description of each individual columns can be found on *<brfss_codebook.html>*. 
I used this webpage to get a better understanding of dataset and select the appropriate data from it for my project.

* * *

## Part 2: Research questions

**Research quesion 1:**
Which state has the highest asthma cases in ratio/percentage?
Is the ranking in asthmatic cases consistent over recent years?
Possible Reasons for high rate of asthmatic patients.

Asthma is relatively common illness. I wish to find out which states top the list in asthma cases
in terms of sample ration in dataset. We can then do more research on why higher rate of asthma cases
in the area.

**Research quesion 2:**
Does being overweight results in higher probability in heart attack?
Does obese people exercise less frequently?

Being overweight is a health problem brough by ourselves. I wish the dataset would tell us overweight
is a health problem we can solve by doing exercise regularly.

**Research quesion 3:**
Does low income level affect people from getting healthcare covreage?
Does low income level restrict people from getting proper medications?

I believe poor people should also have adequate form of healthcare coerage.
I would like to see from plots if there is trend that people from lower income group
have lower healthecare coerage and wish to highloight it from my findings in the dataset.

* * *

## Part 3: Exploratory data analysis

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button (green button with orange arrow) above. Make sure that your code is visible
in the project you submit. Delete this note when before you submit your work.

**Research quesion 1:**
Which state has the highest asthma cases in ratio/percentage?
Is the ranking in asthmatic cases consistent over recent years?
Possible Reasons for high rate of asthmatic patients.

```{r}
# To find out which state has the highest asthma rate ratio.

# asthma3: Ever Told Had Asthma

ggplot(data=brfss2013_truncate, aes(x = X_state, fill = asthma3)) +
  geom_bar() +
  labs(title = "asthma3: Ever Told Had Asthma") +
  theme(axis.text.x = element_text(angle=90))

# As each state has different sample sizes, this plot is hard to tell which state fare better than
# others in terms of asthmaic patients
# A better appproach is to calculate ratio of Yes/No for each state.

```


```{r}
# List the state name in column 'X_state'
unique(brfss2013_truncate$X_state) 

# Data cleaning:
# Need to exclude the '0' and '80' from X_state column, which has no meaning in state name

df <- brfss2013_truncate %>% filter(!X_state %in% c(0,80))

dim(df) # Check 2 unwanted rows are deleted



```


```{r}
# List again the state name in column 'X_state' after remove unwanted rows
unique(df$X_state)

```


```{r}
# asthma3: Ever Told Had Asthma

# Count the number of "Yes", "No", & NA in each state.

asthmaYN <- df %>% 
  select(X_state, asthma3) %>% 
  group_by(X_state, asthma3) %>% 
  count() #  <==> summarise(n = sum(asthma3 %in% c("Yes", "No", NA)))

asthmaYN
```

```{r}

# Consolidate the "Yes" & "No" into one column each.
# Also create a new column of ratio of YEs/No for each state. This will create a better view when plot.

asthmaYes <- asthmaYN %>% filter(asthma3 == "Yes") %>% rename(stateYes = n)
asthmaYes <- asthmaYes[c("X_state","stateYes")] # Remove column "asthma3"

asthmaNo <- asthmaYN %>% filter(asthma3 == "No") %>% rename(stateNo = n)

asthmaYesNo <- cbind(asthmaYes,
                     stateNo = asthmaNo$stateNo,
                     ratioYes = asthmaYes$stateYes/(asthmaYes$stateYes + asthmaNo$stateNo))

head(asthmaYesNo, n=10)
```


```{r}

ggplot(data = asthmaYesNo, aes(x= X_state, y =ratioYes)) +
  geom_point() +
  theme(axis.text.x = element_text(angle=90))

# This chart shows Oregan has the highest ratio in asthmaic patients among all states
```




```{r}
# Conclusion:
# The asthma rate cases ratio is farely consistent over the recent years when compare to Following CDC website.
# https://www.cdc.gov/asthma/most_recent_data_states.htm

# Oregan, although not the highest percentage of asthma cases, but is among the highest in dicated in the website.

# I guess it could be due to the devastating wild forest fire (Sep-2020) that covered a vast area in Oregan.
# Or it could be due to wide coverage of pollens from flower plants in Oregan that triggered more asthmatic cases. 
# There could be other overlooked or unmentioned reasons though for high asthma rate.


```



**Research quesion 2:**
Does being overweight results in higher probability in heart attack?
Does obese people exercise less frequently?
  
  
```{r}
#X_bmi5cat: Computed Body Mass Index Categories

#Value          Value Label
# 1     Underweight [*_bmi5* >= 1200 and < 1850]
# 2     Normal Weight [*_bmi5* >= 1850 and < 2500]
# 3     Overweight [*_bmi5* >= 2500 and < 3000]
# 4     Obese [*_bmi5* >= 3000 and < 9999]
# 5     [Missing] [*_bmi5* coded missing]

df %>% group_by(X_bmi5cat) %>% count()


```

```{r}
# cvdinfr4: Ever Diagnosed With Heart Attack

df %>% group_by(cvdinfr4) %>% count()

```


```{r}

ggplot(data=df, aes(x = X_bmi5cat, fill = cvdinfr4)) +
  geom_bar() +
  labs(title='Number of heart attack/no heart attack for each weight category')

```
The chart shows more number of "Yes" under the 'Normal weight', 'Overweight', and 'Obese' category.
In terms of ratio of yes/no per category, it may tell a different story. 
Lets create a plot in terms of ratio.

```{r}

df2 <-df %>%
  group_by(X_bmi5cat,cvdinfr4) %>% count()

df2

```


```{r}
hAttack <- df2 %>% filter(cvdinfr4 == "Yes") %>% rename(yesAttack = n)
hAttack <- hAttack[,-c(2)]

nohAttack <-df2 %>% filter(cvdinfr4 == "No") %>% rename(noAttack = n)

names(hAttack)  
hAttackYesNo <- cbind(hAttack,
                 noAttack = nohAttack$noAttack,
                 ratioYesNo = hAttack$yesAttack/(hAttack$yesAttack + nohAttack$noAttack) )
hAttackYesNo

```





```{r}
ggplot(data=hAttackYesNo, aes(x = X_bmi5cat, y=ratioYesNo)) +
  geom_point(aes(size=ratioYesNo)) +
  labs(title='Rate of having heart attack for each weight category')


```
Above plot shows otherwise when compared to above bar plot. The 'Normal weight' shows lowest rate 
of having heart attack.
'Obese' has the highest ratio/percentage rating of heart attack.
Suprisingly, 'Underweight' and 'Overweight' have similiar ratio/percentage ratings in terms of heart attack probability!


What is the trend of people exercising against different obesity levels ? Lets check out the 'exerany2' variable against the 'X_bmi5cat' variable.

```{r}
# X_bmi5cat: Computed Body Mass Index Categories
# exerany2: Exercise In Past 30 Days

ggplot(data=df, aes(x = X_bmi5cat, fill = exerany2)) +
  geom_bar() +
  labs(title="X_bmi5cat: Exercise In Past 30 Days")

```
As the count figure is high into the 150K range, we might not be able to see a good picture of the plot. Lets do another plot in terms of ratio of "Yes' and "No". Some rearranging and consolidating work below on 'exerany2' yes/no data.



```{r}
df2b <-df %>%
  group_by(X_bmi5cat,exerany2) %>%
  count()

df2b

```


```{r}
# Consolidate all the "Yes" into one column, and all the "No" in another column.

yesExer <- df2b %>% filter(exerany2 == "Yes") %>% rename(yesExercise = n)
yesExer <- yesExer[,-c(2)] # Remove unwanted column

noExer <-df2b %>% filter(exerany2 == "No") %>% rename(noExercise = n)


exerYesNo <- cbind(yesExer,
                 noExercise = noExer$noExercise,
                 ratioYesNo = yesExer$yesExercise/(yesExer$yesExercise + noExer$noExercise))
exerYesNo


```



```{r}
ggplot(data=exerYesNo, aes(x = X_bmi5cat, y=ratioYesNo)) +
  geom_point(aes(size=ratioYesNo)) +
  labs(title='Exercise for past 30 days for each weight category')


```

Above plot shows as people get more obese, the lesser frequency they do exercises.
As for the 'Underweight'category, I suspect a number these group of people might have some health
problems, resulting in low exercise take-up rate.

Conclusion:
Overweight has close relationship with heart attack as shown on plotted chart. The more obese, 
the higher chance of getting heart attack.
Conversely, overweight people tend not to do exercises regularly. And the more overweight,
the less likely these group of obese people do exercises regularly.




**Research quesion 3:**
Does low income level affect people from getting healthcare covreage?
Does low income level restrict people from getting proper medications?
```{r}

df %>% group_by(income2) %>% summarise(n())

```


```{r}

# hlthpln1: Have Any Health Care Coverage
ggplot(data=df, aes(x = income2, fill = hlthpln1)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle=90)) +
  labs(title = "hlthpln1: Have Any Health Care Coverage")

```
This plot shows lower income level does affect people from getting healthcare coverage.
```{r}

# medcost: Could Not See Dr. Because Of Cost

ggplot(data=df, aes(x = income2, fill = medcost)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle=90)) +
  labs(title = "medcost: Could Not See Dr. Because Of Cost")

```
This plot shows lower income level also caused people to access medication.

Conclusion:
Lower income level does affect people from getting healthcare coverage as shown on plotted bar chart.
Lower income people also have issues seeing doctors.


